---
title: "Project in Statistics - Stochastic Differential Equations"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Packages
```{r, message=FALSE}
library(curl)
library(tidyverse)
library(zoo)
library(expm)
```


# SDE

## Data
```{r, include=FALSE}
excel_url <- "https://www.iceandclimate.nbi.ku.dk/data/GICC05modelext_GRIP_and_GISP2_and_resampled_data_series_Seierstad_et_al._2014_version_10Dec2014-2.xlsx"

data_file <- file.path("/Users/valdemarcronfeldschaaf/Desktop/Matematik/Kandidat/25:26/Blok 4/Project in Statistics", "Ca2data.csv")

res <- httr::GET(excel_url)
excel_file <- tempfile(fileext = ".xlsx")
writeBin(httr::content(res, as = "raw"), excel_file)
Ca2data <- readxl::read_excel(excel_file, sheet = 3,
range = "A51:I12278") %>%
dplyr::slice(-1) %>%
dplyr::select(1, ncol(.) - 1, ncol(.)) %>%
dplyr::mutate_all(as.numeric) %>%
dplyr::rename(age = 1, d18O = 2, Ca2 = 3)
unlink(excel_file)
readr::write_csv(Ca2data, data_file)
```

```{r}
head(Ca2data)
```
### ... data
```{r}
Ca2_final <- Ca2data %>%
  select(-d18O) %>%
  filter(age >= 30000, age <= 80000) %>%
  mutate(Ca2 = na.approx(Ca2)) %>%
  group_by(age) %>%
  summarise(Ca2 = mean(Ca2), .groups = "drop") %>%
  mutate(Ca2 = -log(Ca2)) %>%
  mutate(X = Ca2 - mean(Ca2)) %>%
  select(-Ca2) %>%
  mutate(age = age/1000) %>%
  mutate(V = (lead(X) - X)/0.02) %>% 
  arrange(desc(age))

head(Ca2_final)
```

```{r}
ggplot(Ca2_final, aes(x = age, y = X)) +
  geom_line(color = "black") +
  scale_x_reverse() +
  labs(title = "Plot of Data",
       x = "Age (kyr BP)", 
       y = "X_t") +
  theme_minimal()
```


## Implementing our functions

Our functions should depend on data and theta, with theta being

\begin{align*}
\theta=(\eta,a,b,c,d,\alpha,\beta,\gamma)
\end{align*}

\begin{align*}
\eta &\geq 0 \\
a &< 0 \\
\beta^2 &< 4 \alpha \gamma \\
0 < \alpha &< 2 \eta
\end{align*}


### theta
```{r}
# theta = (eta,a,b,c,d,alpha,beta,gamma)
theta <- c(1,2,3,4,5,6,7,8)
theta_test <- c(67, -171, 3, 340, 31, 86, 27, 434)
theta_test2 <- c(25, -171, 3, 340, 31,  36, 28, 430)
theta_test3 <- c(50, -200, 10, 100, 10,  30, 5, 1000)
# eta >= 0
# a < 0
# beta^2 < 4*alpha*gamma
# 0 < alpha < 2*eta

```

### Python SS pseudo log-likelihood translated to R
This is a direct R translation of the Python function `SS_neg_log_lik` from
`student_kramers/likelihoods.py`. The Python function returns a negative
log-likelihood, so `SS_loglikelihood_R()` below changes the sign to match the
convention used in `pseudo_loglikelihood()` above.

```{r}
SS_precompute_matrices_R <- function(theta, drift_const, h) {
  eta <- theta[1]
  alpha <- theta[6]
  beta <- theta[7]
  gamma <- theta[8]

  A_ss <- matrix(c(0, drift_const,
                   1, -eta),
                 nrow = 2, ncol = 2)
  I2 <- diag(2)
  A_kron <- kronecker(A_ss, I2) + kronecker(I2, A_ss)

  alpha_mat <- matrix(0, nrow = 4, ncol = 4)
  alpha_mat[4, 4] <- alpha

  block_exp <- function(TL, TR, BR) {
    n <- nrow(TL)
    P <- matrix(0, nrow = 2 * n, ncol = 2 * n)
    P[1:n, 1:n] <- TL
    P[1:n, (n + 1):(2 * n)] <- TR
    P[(n + 1):(2 * n), (n + 1):(2 * n)] <- BR
    expm(h * P)[1:n, (n + 1):(2 * n), drop = FALSE]
  }

  I1 <- block_exp(A_kron + alpha_mat, alpha_mat, A_kron)
  I3 <- block_exp(
    A_kron + alpha_mat,
    alpha_mat,
    kronecker(A_ss, I2) + diag(4)
  )
  I5 <- block_exp(A_kron + alpha_mat, diag(4), matrix(0, 4, 4))
  I5_G5 <- as.numeric(I5 %*% c(0, 0, 0, gamma))

  P4 <- matrix(0, nrow = 6, ncol = 6)
  P4[1:4, 1:4] <- A_kron + alpha_mat
  P4[4, 6] <- beta
  P4[5:6, 5:6] <- A_ss
  I4 <- expm(h * P4)[1:4, 5:6, drop = FALSE]

  list(
    A = A_ss,
    exp_Ah = expm(h * A_ss),
    I1 = I1,
    I3 = I3,
    I4 = I4,
    I5_G5 = I5_G5
  )
}

SS_step_omega_R <- function(Y_mid, kappa, mats) {
  x0 <- Y_mid[1] - kappa
  x1 <- Y_mid[2]

  x_outer <- c(x0 * x0, x0 * x1, x1 * x0, x1 * x1)
  xxs <- c(x0 * kappa, 0, x1 * kappa, 0)

  omega_vec <- as.numeric(
    mats$I1 %*% x_outer +
      2 * (mats$I3 %*% xxs) +
      mats$I4 %*% c(x0, x1) +
      mats$I5_G5
  )

  matrix(omega_vec, nrow = 2, ncol = 2, byrow = TRUE)
}

SS_f_step_R <- function(Y, h_step, theta, drift_const, kappa) {
  a <- theta[2]
  b <- theta[3]
  c <- theta[4]
  d <- theta[5]

  X <- Y[1]
  V <- Y[2]

  c(
    X,
    V + h_step * (a * X^3 + b * X^2 + c * X + d -
                    drift_const * (X - kappa))
  )
}

SS_step_R <- function(Y_old, Y_new, h, theta, drift_const,
                      kappa_pos, kappa_neg, mats) {
  kappa <- if (Y_old[1] > 0) kappa_pos else kappa_neg

  f_new <- SS_f_step_R(Y_new, -h / 2, theta, drift_const, kappa)
  Y_mid <- SS_f_step_R(Y_old,  h / 2, theta, drift_const, kappa)

  center <- c(kappa, 0)
  mu <- as.numeric(mats$exp_Ah %*% (Y_mid - center) + center)
  z <- f_new - mu

  Omega <- SS_step_omega_R(Y_mid, kappa, mats)
  omega11 <- Omega[1, 1]
  omega12 <- Omega[1, 2]
  omega21 <- Omega[2, 1]
  omega22 <- Omega[2, 2]

  det_omega <- max(omega11 * omega22 - omega12 * omega21, 1e-8)
  quadratic <- (
    omega22 * z[1]^2 -
      (omega12 + omega21) * z[1] * z[2] +
      omega11 * z[2]^2
  ) / det_omega

  0.5 * (log(det_omega) + quadratic)
}

SS_neg_log_lik_R <- function(theta, data, h) {
  X <- as.matrix(data[, c("X", "V")])
  X <- X[complete.cases(X), , drop = FALSE]

  if (nrow(X) < 2) {
    stop("Need at least two complete observations.")
  }

  data_old <- X[-nrow(X), , drop = FALSE]
  data_new <- X[-1, , drop = FALSE]

  a <- theta[2]
  b <- theta[3]
  c <- theta[4]

  m <- mean(data_old[, 1])
  m2 <- mean(data_old[, 1]^2)
  var_x <- var(data_old[, 1])

  drift_const <- 3 * a * m2 + 2 * b * m + c
  root <- sqrt(max((3 * a * m + b)^2 + 9 * a^2 * var_x, 0))
  kappa_pos <- (-b - root) / (3 * a)
  kappa_neg <- (-b + root) / (3 * a)

  mats <- SS_precompute_matrices_R(theta, drift_const, h)

  nlls <- vapply(
    seq_len(nrow(data_old)),
    function(i) {
      SS_step_R(
        Y_old = data_old[i, ],
        Y_new = data_new[i, ],
        h = h,
        theta = theta,
        drift_const = drift_const,
        kappa_pos = kappa_pos,
        kappa_neg = kappa_neg,
        mats = mats
      )
    },
    numeric(1)
  )

  sum(nlls)
}

SS_loglikelihood_R <- function(data, h, theta) {
  -SS_neg_log_lik_R(theta = theta, data = data, h = h)
}

# Same sign convention as pseudo_loglikelihood():
SS_loglikelihood_R(Ca2_final, h = 0.02, theta_test)
```

### Python SS optimizer translated to R
This mirrors the repository's `run_estimator_lbfgs()` workflow: build a loss
function from the SS negative log-likelihood, minimize it, and return the
estimated parameter vector.

```{r}
run_estimator_lbfgs_R <- function(params_init, loss_fn,
                                  maxiter = 1000, tol = 1e-5,
                                  verbose = TRUE) {
  objective <- function(params) {
    value <- tryCatch(
      loss_fn(params),
      error = function(e) Inf,
      warning = function(w) suppressWarnings(loss_fn(params))
    )

    if (!is.finite(value)) {
      return(1e100)
    }

    value
  }

  fit <- optim(
    par = params_init,
    fn = objective,
    method = "L-BFGS-B",
    control = list(
      maxit = maxiter,
      pgtol = tol,
      parscale = pmax(abs(params_init), 1),
      trace = if (verbose) 1 else 0
    )
  )

  if (verbose) {
    cat("--- L-BFGS-B ---\n")
    cat("  final NLL:", fit$value, "\n")
    cat("  convergence:", fit$convergence, "\n")
  }

  list(
    params_hat = fit$par,
    final_loss = fit$value,
    convergence = fit$convergence,
    converged = fit$convergence == 0,
    message = fit$message,
    counts = fit$counts,
    optim = fit
  )
}

make_SS_loss_fn_R <- function(data, h) {
  function(theta) {
    SS_neg_log_lik_R(theta = theta, data = data, h = h)
  }
}

estimate_SS_theta_R <- function(data, h, theta_start,
                                maxiter = 1000, tol = 1e-5,
                                verbose = TRUE) {
  loss_fn <- make_SS_loss_fn_R(data, h)
  fit <- run_estimator_lbfgs_R(
    params_init = theta_start,
    loss_fn = loss_fn,
    maxiter = maxiter,
    tol = tol,
    verbose = verbose
  )

  names(fit$params_hat) <- c("eta", "a", "b", "c", "d", "alpha", "beta", "gamma")
  fit$loglikelihood <- -fit$final_loss
  fit
}

SS_theta_checks_R <- function(theta) {
  eta <- theta[1]
  a <- theta[2]
  alpha <- theta[6]
  beta <- theta[7]
  gamma <- theta[8]

  data.frame(
    condition = c(
      "all finite",
      "eta > 0",
      "a < 0",
      "alpha > 0",
      "alpha < 2 * eta",
      "beta^2 < 4 * alpha * gamma"
    ),
    value = c(
      all(is.finite(theta)),
      eta,
      a,
      alpha,
      alpha - 2 * eta,
      beta^2 - 4 * alpha * gamma
    ),
    ok = c(
      all(is.finite(theta)),
      eta > 0,
      a < 0,
      alpha > 0,
      alpha < 2 * eta,
      beta^2 < 4 * alpha * gamma
    )
  )
}

# Example:
fit_SS <- estimate_SS_theta_R(Ca2_final, h = 0.02, theta_start = theta_test)
fit_SS$params_hat
fit_SS$loglikelihood
SS_theta_checks_R(fit_SS$params_hat)
```

### Simulate
```{r}
simulation_M <- function(N, h, theta, start = c(1.5,0), seed = 67) {
  
  eta <- theta[1]
  a <- theta[2]
  b <- theta[3]
  c <- theta[4]
  d <- theta[5]
  alpha <- theta[6]
  beta <- theta[7]
  gamma <- theta[8]
  
  X <- numeric(N)
  V <- numeric(N)
  
  X[1] <- start[1]
  V[1] <- start[2]
  W <- rnorm(N, 0, sqrt(h))
  for (i in 2:N) {
    X[i] <- X[i-1] + h * V[i-1]
    V[i] <- V[i-1] + h * (-eta * V[i-1] + a * X[i-1]^3 + b * X[i-1]^2 + c * X[i-1] + d) +
      sqrt(alpha * V[i-1]^2 + beta * V[i-1] + gamma) * W[i] + 0.5 * (alpha * V[i-1] + beta / 2) * (W[i]^2 - h)
  }
  return(data.frame(
    time = seq(0, by = h, length.out = N),
    X = X,
    V = V
  ))
}
```

```{r}
sim_M <- simulation_M(
  N = 1000000,
  h = 0.0001,
  theta = c(30, -125, 40, 150, -20, 20, -8, 1280.8),
  start = c(1.5,0),
  seed = 67
)
sim_M
plot(sim_M$time, sim_M$X, type = "l", xlab = "time", ylab = "X")
plot(sim_M$time, sim_M$V, type = "l", xlab = "time", ylab = "V")
```
```{r}
theta0 <- c(
  eta   = 30,
  a     = -125,
  b     = 40,
  c     = 150,
  d     = -20,
  alpha = 20,
  beta  = -8,
  gamma = 1280
)

T <- 10
h <- 0.001
N <- T / h + 1

n_sims <- 200

sim_list <- lapply(1:n_sims, function(j) {
  dat <- simulation_M(
    N = N,
    h = h,
    theta = theta0,
    start = c(1.5, 0),
    seed = 67 + j
  )

  dat$dataset <- j
  dat
})

sim_data <- do.call(rbind, sim_list)
sim_data
```

```{r}
sim_data_1 <- sim_data %>%
  filter(dataset == 1) %>%
  select(time, X, V)

SS_neg_log_lik_R(theta = theta0, data = sim_data_1, h = h)
SS_loglikelihood_R(data = sim_data_1, h = h, theta = theta0)

fit_sim_data_1 <- estimate_SS_theta_R(
  data = sim_data_1,
  h = h,
  theta_start = theta_test3,
  maxiter = 100,
  verbose = TRUE
)

fit_sim_data_1$params_hat
fit_sim_data_1$final_loss
fit_sim_data_1$loglikelihood
SS_theta_checks_R(fit_sim_data_1$params_hat)
```

```{r}
datasets <- split(sim_data, sim_data$dataset)
```


```{r}
estimate_one_dataset_SS <- function(dataset_id, data_i, h, theta_start,
                                    theta_true, maxiter = 100) {
  data_i <- data_i %>%
    select(time, X, V)

  fit_i <- tryCatch(
    estimate_SS_theta_R(
      data = data_i,
      h = h,
      theta_start = theta_start,
      maxiter = maxiter,
      verbose = FALSE
    ),
    error = function(e) NULL
  )

  if (is.null(fit_i)) {
    return(data.frame(
      dataset = dataset_id,
      parameter = names(theta_true),
      theta_true = as.numeric(theta_true),
      theta_hat = NA_real_,
      relative_error = NA_real_,
      final_loss = NA_real_,
      loglikelihood = NA_real_,
      converged = FALSE,
      theta_valid = FALSE,
      failed = TRUE
    ))
  }

  checks_i <- SS_theta_checks_R(fit_i$params_hat)
  theta_valid_i <- all(checks_i$ok)

  data.frame(
    dataset = dataset_id,
    parameter = names(theta_true),
    theta_true = as.numeric(theta_true),
    theta_hat = as.numeric(fit_i$params_hat),
    relative_error = (as.numeric(fit_i$params_hat) - as.numeric(theta_true)) /
      as.numeric(theta_true),
    final_loss = fit_i$final_loss,
    loglikelihood = fit_i$loglikelihood,
    converged = fit_i$converged,
    theta_valid = theta_valid_i,
    failed = !isTRUE(fit_i$converged) ||
      !theta_valid_i ||
      any(!is.finite(fit_i$params_hat))
  )
}

estimate_all_datasets_SS <- function(sim_data, h, theta_start, theta_true,
                                     maxiter = 100) {
  datasets <- split(sim_data, sim_data$dataset)

  fits <- lapply(seq_along(datasets), function(i) {
    estimate_one_dataset_SS(
      dataset_id = as.integer(names(datasets)[i]),
      data_i = datasets[[i]],
      h = h,
      theta_start = theta_start,
      theta_true = theta_true,
      maxiter = maxiter
    )
  })

  bind_rows(fits)
}

params_SS <- estimate_all_datasets_SS(
  sim_data = sim_data,
  h = h,
  theta_start = theta_test3,
  theta_true = theta0,
  maxiter = 100
)

params_SS
```

```{r}
params_SS_summary <- params_SS %>%
  group_by(parameter) %>%
  summarise(
    n_datasets = n_distinct(dataset),
    n_failed = sum(failed),
    mean_theta_hat = mean(theta_hat, na.rm = TRUE),
    median_theta_hat = median(theta_hat, na.rm = TRUE),
    mean_relative_error = mean(relative_error, na.rm = TRUE),
    sd_relative_error = sd(relative_error, na.rm = TRUE),
    .groups = "drop"
  )

params_SS_summary
```
```{r}
readr::write_csv(params_SS, "params_SS.csv")
```
































