---
title: "Project in Statistics - Stochastic Differential Equations"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Packages
```{r, message=FALSE}
library(curl)
library(tidyverse)
library(zoo)
library(expm)
```


# SDE

## Data
```{r, include=FALSE}
excel_url <- "https://www.iceandclimate.nbi.ku.dk/data/GICC05modelext_GRIP_and_GISP2_and_resampled_data_series_Seierstad_et_al._2014_version_10Dec2014-2.xlsx"

data_file <- file.path("/Users/valdemarcronfeldschaaf/Desktop/Matematik/Kandidat/25:26/Blok 4/Project in Statistics", "Ca2data.csv")

res <- httr::GET(excel_url)
excel_file <- tempfile(fileext = ".xlsx")
writeBin(httr::content(res, as = "raw"), excel_file)
Ca2data <- readxl::read_excel(excel_file, sheet = 3,
range = "A51:I12278") %>%
dplyr::slice(-1) %>%
dplyr::select(1, ncol(.) - 1, ncol(.)) %>%
dplyr::mutate_all(as.numeric) %>%
dplyr::rename(age = 1, d18O = 2, Ca2 = 3)
unlink(excel_file)
readr::write_csv(Ca2data, data_file)
```

```{r}
head(Ca2data)
```
### ... data
```{r}
Ca2_final <- Ca2data %>%
  select(-d18O) %>%
  filter(age >= 30000, age <= 80000) %>%
  mutate(Ca2 = na.approx(Ca2)) %>%
  group_by(age) %>%
  summarise(Ca2 = mean(Ca2), .groups = "drop") %>%
  mutate(Ca2 = -log(Ca2)) %>%
  mutate(X = Ca2 - mean(Ca2)) %>%
  select(-Ca2) %>%
  mutate(age = age/1000) %>%
  mutate(V = (lead(X) - X)/0.02)

head(Ca2_final)
```

## Implementing our functions

Our functions should depend on data and theta, with theta being

\begin{align*}
\theta=(\eta,a,b,c,d,\alpha,\beta,\gamma)
\end{align*}

\begin{align*}
\eta &\geq 0 \\
a &< 0 \\
\beta^2 &< 4 \alpha \gamma \\
0 < \alpha &< 2 \eta
\end{align*}


### theta
```{r}
# theta = (eta,a,b,c,d,alpha,beta,gamma)
theta <- c(1,2,3,4,5,6,7,8)
theta_test <- c(67, -171, 3, 340, 31, 86, 27, 434)
# eta >= 0
# a < 0
# beta^2 < 4*alpha*gamma
# 0 < alpha < 2*eta

```

### Python SS pseudo log-likelihood translated to R
This is a direct R translation of the Python function `SS_neg_log_lik` from
`student_kramers/likelihoods.py`. The Python function returns a negative
log-likelihood, so `SS_loglikelihood_R()` below changes the sign to match the
convention used in `pseudo_loglikelihood()` above.

```{r}
SS_precompute_matrices_R <- function(theta, drift_const, h) {
  eta <- theta[1]
  alpha <- theta[6]
  beta <- theta[7]
  gamma <- theta[8]

  A_ss <- matrix(c(0, drift_const,
                   1, -eta),
                 nrow = 2, ncol = 2)
  I2 <- diag(2)
  A_kron <- kronecker(A_ss, I2) + kronecker(I2, A_ss)

  alpha_mat <- matrix(0, nrow = 4, ncol = 4)
  alpha_mat[4, 4] <- alpha

  block_exp <- function(TL, TR, BR) {
    n <- nrow(TL)
    P <- matrix(0, nrow = 2 * n, ncol = 2 * n)
    P[1:n, 1:n] <- TL
    P[1:n, (n + 1):(2 * n)] <- TR
    P[(n + 1):(2 * n), (n + 1):(2 * n)] <- BR
    expm(h * P)[1:n, (n + 1):(2 * n), drop = FALSE]
  }

  I1 <- block_exp(A_kron + alpha_mat, alpha_mat, A_kron)
  I3 <- block_exp(
    A_kron + alpha_mat,
    alpha_mat,
    kronecker(A_ss, I2) + diag(4)
  )
  I5 <- block_exp(A_kron + alpha_mat, diag(4), matrix(0, 4, 4))
  I5_G5 <- as.numeric(I5 %*% c(0, 0, 0, gamma))

  P4 <- matrix(0, nrow = 6, ncol = 6)
  P4[1:4, 1:4] <- A_kron + alpha_mat
  P4[4, 6] <- beta
  P4[5:6, 5:6] <- A_ss
  I4 <- expm(h * P4)[1:4, 5:6, drop = FALSE]

  list(
    A = A_ss,
    exp_Ah = expm(h * A_ss),
    I1 = I1,
    I3 = I3,
    I4 = I4,
    I5_G5 = I5_G5
  )
}

SS_step_omega_R <- function(Y_mid, kappa, mats) {
  x0 <- Y_mid[1] - kappa
  x1 <- Y_mid[2]

  x_outer <- c(x0 * x0, x0 * x1, x1 * x0, x1 * x1)
  xxs <- c(x0 * kappa, 0, x1 * kappa, 0)

  omega_vec <- as.numeric(
    mats$I1 %*% x_outer +
      2 * (mats$I3 %*% xxs) +
      mats$I4 %*% c(x0, x1) +
      mats$I5_G5
  )

  matrix(omega_vec, nrow = 2, ncol = 2, byrow = TRUE)
}

SS_f_step_R <- function(Y, h_step, theta, drift_const, kappa) {
  a <- theta[2]
  b <- theta[3]
  c <- theta[4]
  d <- theta[5]

  X <- Y[1]
  V <- Y[2]

  c(
    X,
    V + h_step * (a * X^3 + b * X^2 + c * X + d -
                    drift_const * (X - kappa))
  )
}

SS_step_R <- function(Y_old, Y_new, h, theta, drift_const,
                      kappa_pos, kappa_neg, mats, mats_15h) {
  kappa <- if (Y_old[1] > 0) kappa_pos else kappa_neg

  f_new <- SS_f_step_R(Y_new, -h / 2, theta, drift_const, kappa)
  Y_mid <- SS_f_step_R(Y_old,  h / 2, theta, drift_const, kappa)

  center <- c(kappa, 0)
  mu <- as.numeric(mats$exp_Ah %*% (Y_mid - center) + center)
  z <- f_new - mu

  omega22 <- SS_step_omega_R(Y_mid, kappa, mats)[2,2]
  #omega11 <- Omega[1, 1] # needs to be inverted
  #omega12 <- Omega[1, 2]
  #omega21 <- Omega[2, 1]
  #omega22 <- Omega[2, 2]
  
  Y_mid1 <- SS_f_step_R(Y_old,  3/2 * h / 2, theta, drift_const, kappa)

  omegacor22 <- SS_step_omega_R(Y_mid1, kappa, mats_15h)[2, 2]
  #omegacor11 <- Omegacor[1, 1]
  #omegacor12 <- Omegacor[1, 2]
  #omegacor21 <- Omegacor[2, 1]
  #omegacor22 <- Omegacor[2, 2]

  #det_omegacor <- max(omegacor11 * omegacor22 - omegacor12 * omegacor21, 1e-8)
  quadratic <- (
    #omega22 * z[1]^2 -
      #(omega12 + omega21) * z[1] * z[2] +
     z[2]^2 / omega22 
  ) 

  (2/3 * log(omegacor22) + quadratic)
}

SS_neg_log_lik_R <- function(theta, data, h) {
  X <- as.matrix(data[, c("X", "V")])
  X <- X[complete.cases(X), , drop = FALSE]
  
  if (nrow(X) < 2) {
    stop("Need at least two complete observations.")
  }

  data_old <- X[-nrow(X), , drop = FALSE]
  data_new <- X[-1, , drop = FALSE]

  a <- theta[2]
  b <- theta[3]
  c <- theta[4]

  m <- mean(data_old[, 1])
  m2 <- mean(data_old[, 1]^2)
  var_x <- var(data_old[, 1])

  drift_const <- 3 * a * m2 + 2 * b * m + c
  root <- sqrt(max((3 * a * m + b)^2 + 9 * a^2 * var_x, 0))
  kappa_pos <- (-b - root) / (3 * a)
  kappa_neg <- (-b + root) / (3 * a)

  mats <- SS_precompute_matrices_R(theta, drift_const, h)
  mats_15h <- SS_precompute_matrices_R(theta, drift_const, 1.5 * h)


  nlls <- vapply(
    seq_len(nrow(data_old)),
    function(i) {
      SS_step_R(
        Y_old = data_old[i, ],
        Y_new = data_new[i, ],
        h = h,
        theta = theta,
        drift_const = drift_const,
        kappa_pos = kappa_pos,
        kappa_neg = kappa_neg,
        mats = mats,
        mats_15h = mats_15h
      )
    },
    numeric(1)
  )

  sum(nlls)
}

SS_loglikelihood_R <- function(data, h, theta) {
  -SS_neg_log_lik_R(theta = theta, data = data, h = h)
}

# Same sign convention as pseudo_loglikelihood():
SS_loglikelihood_R(Ca2_final, h = 0.02, theta_test)
```

```{r}
Ca2data_new <- cbind(Ca2_final$X, Ca2_final$V)
head(Ca2data)
```


### Python SS optimizer translated to R
This mirrors the repository's `run_estimator_lbfgs()` workflow: build a loss
function from the SS negative log-likelihood, minimize it, and return the
estimated parameter vector.

```{r}
run_estimator_lbfgs_R <- function(params_init, loss_fn,
                                  maxiter = 1000, tol = 1e-5,
                                  verbose = TRUE) {
  objective <- function(params) {
    value <- tryCatch(
      loss_fn(params),
      error = function(e) Inf,
      warning = function(w) suppressWarnings(loss_fn(params))
    )

    if (!is.finite(value)) {
      return(1e100)
    }

    value
  }

  fit <- optim(
    par = params_init,
    fn = objective,
    method = "L-BFGS-B",
    control = list(
      maxit = maxiter,
      pgtol = tol,
      parscale = pmax(abs(params_init), 1),
      trace = if (verbose) 1 else 0
    )
  )

  if (verbose) {
    cat("--- L-BFGS-B ---\n")
    cat("  final NLL:", fit$value, "\n")
    cat("  convergence:", fit$convergence, "\n")
  }

  list(
    params_hat = fit$par,
    final_loss = fit$value,
    convergence = fit$convergence,
    converged = fit$convergence == 0,
    message = fit$message,
    counts = fit$counts,
    optim = fit
  )
}

make_SS_loss_fn_R <- function(data, h) {
  function(theta) {
    SS_neg_log_lik_R(theta = theta, data = data, h = h)
  }
}

estimate_SS_theta_R <- function(data, h, theta_start,
                                maxiter = 1000, tol = 1e-5,
                                verbose = TRUE) {
  loss_fn <- make_SS_loss_fn_R(data, h)
  fit <- run_estimator_lbfgs_R(
    params_init = theta_start,
    loss_fn = loss_fn,
    maxiter = maxiter,
    tol = tol,
    verbose = verbose
  )

  names(fit$params_hat) <- c("eta", "a", "b", "c", "d", "alpha", "beta", "gamma")
  fit$loglikelihood <- -fit$final_loss
  fit
}

SS_theta_checks_R <- function(theta) {
  eta <- theta[1]
  a <- theta[2]
  alpha <- theta[6]
  beta <- theta[7]
  gamma <- theta[8]

  data.frame(
    condition = c(
      "all finite",
      "eta > 0",
      "a < 0",
      "alpha > 0",
      "alpha < 2 * eta",
      "beta^2 < 4 * alpha * gamma"
    ),
    value = c(
      all(is.finite(theta)),
      eta,
      a,
      alpha,
      alpha - 2 * eta,
      beta^2 - 4 * alpha * gamma
    ),
    ok = c(
      all(is.finite(theta)),
      eta > 0,
      a < 0,
      alpha > 0,
      alpha < 2 * eta,
      beta^2 < 4 * alpha * gamma
    )
  )
}

# Example:
fit_SS <- estimate_SS_theta_R(Ca2_final, h = 0.02, theta_start = theta_start)
fit_SS$params_hat
fit_SS$loglikelihood
SS_theta_checks_R(fit_SS$params_hat)
```


















