---
title: "Project in Statistics - Stochastic Differential Equations"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Packages
```{r, message=FALSE}
library(curl)
library(tidyverse)
library(zoo)
library(expm)
```


# SDE

## Data
```{r, include=FALSE}
excel_url <- "https://www.iceandclimate.nbi.ku.dk/data/GICC05modelext_GRIP_and_GISP2_and_resampled_data_series_Seierstad_et_al._2014_version_10Dec2014-2.xlsx"

data_file <- file.path("/Users/valdemarcronfeldschaaf/Desktop/Matematik/Kandidat/25:26/Blok 4/Project in Statistics", "Ca2data.csv")

res <- httr::GET(excel_url)
excel_file <- tempfile(fileext = ".xlsx")
writeBin(httr::content(res, as = "raw"), excel_file)
Ca2data <- readxl::read_excel(excel_file, sheet = 3,
range = "A51:I12278") %>%
dplyr::slice(-1) %>%
dplyr::select(1, ncol(.) - 1, ncol(.)) %>%
dplyr::mutate_all(as.numeric) %>%
dplyr::rename(age = 1, d18O = 2, Ca2 = 3)
unlink(excel_file)
readr::write_csv(Ca2data, data_file)
```

```{r}
head(Ca2data)
```
### ... data
```{r}
Ca2_final <- Ca2data %>%
  select(-d18O) %>%
  filter(age >= 30000, age <= 80000) %>%
  mutate(Ca2 = na.approx(Ca2)) %>%
  group_by(age) %>%
  summarise(Ca2 = mean(Ca2), .groups = "drop") %>%
  mutate(Ca2 = -log(Ca2)) %>%
  mutate(X = Ca2 - mean(Ca2)) %>%
  select(-Ca2) %>%
  mutate(age = age/1000) %>%
  mutate(V = (lead(X) - X)/0.02)

head(Ca2_final)
```

## Implementing our functions

Our functions should depend on data and theta, with theta being

\begin{align*}
\theta=(\eta,a,b,c,d,\alpha,\beta,\gamma)
\end{align*}

\begin{align*}
\eta &\geq 0 \\
a &< 0 \\
\beta^2 &< 4 \alpha \gamma \\
0 < \alpha &< 2 \eta
\end{align*}


### theta
```{r}
# theta = (eta,a,b,c,d,alpha,beta,gamma)
theta <- c(1,2,3,4,5,6,7,8)
# eta >= 0
# a < 0
# beta^2 < 4*alpha*gamma
# 0 < alpha < 2*eta

```

### $A$
```{r}
A <- function(theta){
  A <- matrix(c(0, theta[4], 1, -theta[1]), nrow = 2, ncol = 2)
  return(A)
}
```

### $b$
```{r}
b <- function(theta){
  b <- matrix(c(theta[5]/theta[4],0 ), nrow = 2, ncol = 1)
  return(b)
}
```

### $\mu$
```{r}
mu <- function(data,theta){
  mu <- expm(A*h) %*% (data$X-b(theta)) + b(theta)
  return(b)
}
```

### Defining $\breve{\alpha}$ and $\breve{\beta}$
```{r}
d <- 2

# alpha^{ij} matrices
alpha11 <- matrix(0, d, d)
alpha12 <- matrix(0, d, d)
alpha21 <- matrix(0, d, d)
alpha22 <- matrix(c(0, 0,
                    0, theta[6]),
                  nrow = d, byrow = TRUE)

# beta^{ij} vectors
beta11 <- c(0, 0)
beta12 <- c(0, 0)
beta21 <- c(0, 0)
beta22 <- c(0, theta[7])

# alpha breve: rows are vec(alpha^{ij})^T
# R's as.vector() uses column-wise vectorization, matching mathematical vec()
alpha_breve <- rbind(
  as.vector(alpha11),
  as.vector(alpha12),
  as.vector(alpha21),
  as.vector(alpha22)
)

# beta breve: rows are beta^{ij}^T
beta_breve <- rbind(
  beta11,
  beta12,
  beta21,
  beta22
)

alpha_breve
beta_breve

alpha_breve_fun <- function(theta) {
  alpha22 <- matrix(c(0, 0,
                      0, theta[6]),
                    nrow = d, byrow = TRUE)

  rbind(
    as.vector(alpha11),
    as.vector(alpha12),
    as.vector(alpha21),
    as.vector(alpha22)
  )
}

beta_breve_fun <- function(theta) {
  beta22 <- c(0, theta[7])

  rbind(
    beta11,
    beta12,
    beta21,
    beta22
  )
}
```
### $M_1-M_5$
```{r}
kronecker_sum <- function(A, B) {
  m <- nrow(A)
  n <- nrow(B)

  kronecker(A, diag(n)) + kronecker(diag(m), B)
}
```

```{r}
M1 <- function(theta){
  A_kron <- kronecker_sum(A(theta), A(theta))
  M <- rbind(
    cbind(A_kron + alpha_breve_fun(theta), alpha_breve_fun(theta)),
    cbind(matrix(0,4,4),     A_kron)
  )

  return(M)
}

M2 <- function(theta){
  A_theta <- A(theta)
  A_kron <- kronecker_sum(A_theta, A_theta)
  alpha_breve <- alpha_breve_fun(theta)

  M <- rbind(
    cbind(A_kron + alpha_breve, alpha_breve),
    cbind(matrix(0,4,4),     kronecker_sum(diag(2), A_theta))
  )

  return(M)
}

M3 <- function(theta){
  A_theta <- A(theta)
  A_kron <- kronecker_sum(A_theta, A_theta)
  alpha_breve <- alpha_breve_fun(theta)

  M <- rbind(
    cbind(A_kron + alpha_breve, alpha_breve),
    cbind(matrix(0,4,4),     kronecker_sum(A_theta, diag(2)))
  )

  return(M)
}

M4 <- function(theta){
  A_theta <- A(theta)
  A_kron <- kronecker_sum(A_theta, A_theta)
  alpha_breve <- alpha_breve_fun(theta)
  beta_breve <- beta_breve_fun(theta)

  M <- rbind(
    cbind(A_kron + alpha_breve, beta_breve),
    cbind(matrix(0,2,4), A_theta)
  )

  return(M)
}

M5 <- function(theta){
  A_kron <- kronecker_sum(A(theta), A(theta))
  alpha_breve <- alpha_breve_fun(theta)

  M <- rbind(
    cbind(A_kron + alpha_breve, diag(1,4,4)),
    cbind(matrix(0,4,4), matrix(0,4,4))
  )

  return(M)
}
```


### $I_1-I_5$
```{r}
I1 <- function(h,theta){
  I <- expm(M1(theta)*h)
  return(I[1:4,5:8])
}

I2 <- function(h,theta){
  I <- expm(M2(theta)*h)
  return(I[1:4,5:8])
}

I3 <- function(h,theta){
  I <- expm(M3(theta)*h)
  return(I[1:4,5:8])
}

I4 <- function(h,theta){
  I <- expm(M4(theta)*h)
  return(I[1:4,5:6])
}

I5 <- function(h,theta){
  I <- expm(M5(theta)*h)
  return(I[1:4,5:8])
}
```

### $\Omega$
```{r}
Omega_vec <- function(x, h, theta) {
  x <- as.matrix(x)
  b_theta <- as.numeric(b(theta))
  x_minus_b <- sweep(x, 2, b_theta, "-")

  I1_mat <- I1(h, theta)
  I2_mat <- I2(h, theta)
  #I3_mat <- I3(h, theta)
  I4_mat <- I4(h, theta)
  I5_mat <- I5(h, theta)

  y1 <- x_minus_b[, 1]
  y2 <- x_minus_b[, 2]
  b1 <- b_theta[1]
  b2 <- b_theta[2]

  term1 <- rbind(y1 * y1, y2 * y1, y1 * y2, y2 * y2)
  term2 <- rbind(y1 * b1, y2 * b1, y1 * b2, y2 * b2)
  term3 <- rbind(b1 * y1, b2 * y1, b1 * y2, b2 * y2)
  term4 <- t(x_minus_b)
  term5 <- c(0,0,0,theta[8])
  term5_repeated <- matrix(I5_mat %*% term5, nrow = 4, ncol = nrow(x), byrow = FALSE)

  I1_mat %*% term1 +
    I2_mat %*% term2 +
    I2_mat %*% term3 +
    I4_mat %*% term4 +
    term5_repeated
}

#Omega(c(0, 0), h = 0.02, theta = theta)
```

### Pseudo log-likelihood
```{r}
F_fun <- function(x, theta) {
  x <- as.numeric(x)

  matrix(c(
    x[2],
    -theta[1] * x[2] +
      theta[2] * x[1]^3 +
      theta[3] * x[1]^2 +
      theta[4] * x[1] +
      theta[5]
  ), nrow = 2, ncol = 1)
}

#N_fun <- function(x, theta) {
  #x <- matrix(x, nrow = 2, ncol = 1)
  #F_fun(x, theta) - A(theta) %*% (x - b(theta))
#}

N_mat <- function(x, theta) {
  x <- as.matrix(x)
  b_theta <- as.numeric(b(theta))

  F <- cbind(
    x[, 2],
    -theta[1] * x[, 2] +
      theta[2] * x[, 1]^3 +
      theta[3] * x[, 1]^2 +
      theta[4] * x[, 1] +
      theta[5]
  )

  linear_part <- t(A(theta) %*% t(sweep(x, 2, b_theta, "-")))
  F - linear_part
}

DN_fun <- function(x, theta) {
  x <- as.numeric(x)
  DF <- matrix(c(
    0,
    3 * theta[2] * x[1]^2 + 2 * theta[3] * x[1] + theta[4],
    1,
    -theta[1]
  ), nrow = 2, ncol = 2)

  DF - A(theta)
}

f_half <- function(x, h, theta) {
  x <- matrix(x, nrow = 2, ncol = 1)
  x + (h / 2) * N_fun(x, theta)
}

f_half_inverse <- function(x, h, theta) {
  x <- matrix(x, nrow = 2, ncol = 1)
  x - (h / 2) * N_fun(x, theta)
}

pseudo_loglikelihood <- function(data, h, theta) {
  X <- as.matrix(data[, c("X", "V")])
  X <- X[complete.cases(X), , drop = FALSE]

  #if (nrow(X) < 2) {
    #stop("Need at least two observations.")
  #}

  x_now <- X[-nrow(X), , drop = FALSE]
  x_next <- X[-1, , drop = FALSE]

  N_now <- N_mat(x_now, theta)
  N_next <- N_mat(x_next, theta)

  x_half <- x_now + (h / 2) * N_now
  x_next_back <- x_next - (h / 2) * N_next

  exp_Ah <- expm(A(theta) * h)
  b_theta <- as.numeric(b(theta))
  mean_k <- t(exp_Ah %*% t(sweep(x_half, 2, b_theta, "-")))
  mean_k <- sweep(mean_k, 2, b_theta, "+")

  residual <- x_next_back - mean_k
  omega_vec <- Omega_vec(x_half, h, theta)

  omega11 <- omega_vec[1, ]
  omega21 <- omega_vec[2, ]
  omega12 <- omega_vec[3, ]
  omega22 <- omega_vec[4, ]

  det_omega <- omega11 * omega22 - omega12 * omega21

  #if (any(!is.finite(det_omega)) || any(det_omega <= 0)) {
    #return(-Inf)
  #}

  r1 <- residual[, 1]
  r2 <- residual[, 2]
  quadratic <- (omega22 * r1^2 - (omega12 + omega21) * r1 * r2 + omega11 * r2^2) / det_omega

  #if (any(!is.finite(quadratic))) {
    #return(-Inf)
  #}

  -0.5 * sum(quadratic + log(det_omega))
}

pseudo_loglikelihood(Ca2_final, h = 0.02, theta_test)
```

### Minimize negative pseudo log-likelihood
```{r}
theta_valid <- function(theta) {
  all(theta_checks(theta)$ok)
}

theta_checks <- function(theta) {
  eta <- theta[1]
  a <- theta[2]
  alpha <- theta[6]
  beta <- theta[7]
  gamma <- theta[8]

  data.frame(
    condition = c(
      "all finite",
      "eta > 0",
      "a < 0",
      "alpha > 0",
      "alpha < 2 * eta",
      "beta^2 < 4 * alpha * gamma"
    ),
    value = c(
      all(is.finite(theta)),
      eta,
      a,
      alpha,
      alpha - 2 * eta,
      beta^2 - 4 * alpha * gamma
    ),
    ok = c(
      all(is.finite(theta)),
      eta > 0,
      a < 0,
      alpha > 0,
      alpha < 2 * eta,
      beta^2 < 4 * alpha * gamma
    )
  )
}

theta_to_par <- function(theta) {
  #if (!theta_valid(theta)) {
    #stop("theta_start does not satisfy the parameter constraints.")
  #}

  eta <- theta[1]
  alpha_fraction <- theta[6] / (2 * eta)
  gamma_extra <- theta[8] - theta[7]^2 / (4 * theta[6])

  c(
    log(eta),
    log(-theta[2]),
    theta[3],
    theta[4],
    theta[5],
    qlogis(alpha_fraction),
    theta[7],
    log(gamma_extra)
  )
}

par_to_theta <- function(par) {
  eta <- exp(par[1])
  a <- -exp(par[2])
  alpha <- 2 * eta * plogis(par[6])
  beta <- par[7]
  gamma <- beta^2 / (4 * alpha) + exp(par[8])

  c(eta, a, par[3], par[4], par[5], alpha, beta, gamma)
}

negative_pseudo_loglikelihood <- function(data, h, theta) {
  -pseudo_loglikelihood(data, h, theta)
}

minimize_theta <- function(data, h, theta_start, maxit = 1000) {
  par_start <- theta_to_par(theta_start)

  objective <- function(par) {
    theta_try <- par_to_theta(par)

    value <- tryCatch(
      negative_pseudo_loglikelihood(data, h, theta_try),
      error = function(e) 1e100
    )

    #if (is.finite(value)) value else 1e100
  }

  fit <- optim(
    par = par_start,
    fn = objective,
    method = "L-BFGS-B",
    control = list(
      maxit = maxit,
      parscale = pmax(abs(par_start), 1)
    )
  )

  theta_hat <- par_to_theta(fit$par)

  list(
    theta_hat = theta_hat,
    minimum = fit$value,
    loglikelihood = -fit$value,
    convergence = fit$convergence,
    converged = fit$convergence == 0,
    message = fit$message,
    counts = fit$counts,
    function_evaluations = fit$counts["function"],
    theta_checks = theta_checks(theta_hat),
    theta_valid = theta_valid(theta_hat),
    optim = fit
  )
}
```

```{r, eval=FALSE}
theta_start <- c(50, -200, 10, 100, 10, 30, -5, 1000)
theta_2 <- c(7, -105, 11, 105, 12, 4, -4, 1256)
theta_3 <- c(2, -214, 8, 240, 9, 1, -1, 730)
theta_4 <- c(3, -200, 5, 323, 30, 4, 2, 239)


fit <- minimize_theta(Ca2_final, 0.02, theta_test, maxit = 1000)
fit$theta_hat
fit$minimum
fit$converged
fit$convergence
fit$message
fit$counts
fit$function_evaluations
fit$theta_valid
fit$theta_checks
```

### Direct minimization without reparameterization
```{r}
minimize_theta_direct <- function(data, h, theta_start, maxit = 1000) {
  objective <- function(theta_try) {
    value <- tryCatch(
      negative_pseudo_loglikelihood(data, h, theta_try),
      error = function(e) 1e100
    )

    #if (is.finite(value)) value else 1e100
  }

  fit <- optim(
    par = theta_start,
    fn = objective,
    method = "L-BFGS-B",
    lower = c(0, -Inf, -Inf, -Inf, -Inf, 0, -Inf, 0),
    upper = c(-Inf, 0, -Inf, -Inf, -Inf, -Inf, -Inf, -Inf),
    control = list(
      maxit = maxit,
      parscale = pmax(abs(theta_start), 1)
    )
  )

  theta_hat <- fit$par

  list(
    theta_hat = theta_hat,
    minimum = fit$value,
    loglikelihood = -fit$value,
    convergence = fit$convergence,
    converged = fit$convergence == 0,
    message = fit$message,
    counts = fit$counts,
    function_evaluations = fit$counts["function"],
    theta_checks = theta_checks(theta_hat),
    theta_valid = theta_valid(theta_hat),
    optim = fit
  )
}
```

```{r, eval=FALSE}
theta_new <- c(89.158540, -91.499596, 15.793264, 142.761913, 5.442896, 169.418708, -8.683637, 1262.329676)
fit_direct <- minimize_theta_direct(
  Ca2_final,
  h = 0.02,
  theta_start = theta_test,
  maxit = 10000
)

fit_direct$theta_hat
fit_direct$minimum
fit_direct$converged
fit_direct$convergence
fit_direct$message
fit_direct$counts
fit_direct$function_evaluations

# Test whether the estimated theta satisfies the restrictions
fit_direct$theta_valid
fit_direct$theta_checks
```

## Chapter 4.2 simulation study

### Student Kramers simulation
```{r}
theta_0 <- c(30, -125, 40, 150, -20, 20, -8, 1280.8)
theta_init <- c(50, -200, 10, 100, 10, 30, -5, 1000)
parameter_names <- c("eta", "a", "b", "c", "d", "alpha", "beta", "gamma")

simulate_student_kramers <- function(theta, T = 50, h_sim = 0.0001,
                                     x0 = c(0, 0)) {
  n_steps <- as.integer(T / h_sim)
  X <- numeric(n_steps + 1)
  V <- numeric(n_steps + 1)
  X[1] <- x0[1]
  V[1] <- x0[2]

  eta <- theta[1]
  a <- theta[2]
  b_par <- theta[3]
  c_par <- theta[4]
  d_par <- theta[5]
  alpha <- theta[6]
  beta <- theta[7]
  gamma <- theta[8]

  dW <- rnorm(n_steps, mean = 0, sd = sqrt(h_sim))

  for (i in seq_len(n_steps)) {
    x <- X[i]
    v <- V[i]

    drift_v <- -eta * v + a * x^3 + b_par * x^2 + c_par * x + d_par
    sigma2 <- alpha * v^2 + beta * v + gamma
    sigma <- sqrt(max(sigma2, 0))
    sigma_sigma_prime <- alpha * v + beta / 2

    X[i + 1] <- x + h_sim * v
    V[i + 1] <- v + h_sim * drift_v + sigma * dW[i] +
      0.5 * sigma_sigma_prime * (dW[i]^2 - h_sim)
  }

  data.frame(
    time = seq(0, T, by = h_sim),
    X = X,
    V = V
  )
}

subsample_path <- function(path, h, h_sim = 0.0001) {
  step <- as.integer(round(h / h_sim))
  path[seq(1, nrow(path), by = step), ]
}
```

### Repeated estimation
```{r}
fit_one_simulation <- function(path, h, theta_start = theta_init,
                               maxit = 1000) {
  fit <- tryCatch(
    minimize_theta(path, h = h, theta_start = theta_start, maxit = maxit),
    error = function(e) NULL
  )

  if (is.null(fit)) {
    return(list(
      theta_hat = rep(NA_real_, length(theta_start)),
      minimum = NA_real_,
      loglikelihood = NA_real_,
      converged = FALSE,
      convergence = NA_integer_,
      theta_valid = FALSE,
      failed = TRUE
    ))
  }

  failed <- !isTRUE(fit$converged) ||
    !isTRUE(fit$theta_valid) ||
    any(!is.finite(fit$theta_hat))

  list(
    theta_hat = fit$theta_hat,
    minimum = fit$minimum,
    loglikelihood = fit$loglikelihood,
    converged = fit$converged,
    convergence = fit$convergence,
    theta_valid = fit$theta_valid,
    function_evaluations = fit$function_evaluations,
    failed = failed
  )
}

run_student_kramers_study <- function(n_sim = 20,
                                      h_values = c(0.01, 0.02),
                                      T = 50,
                                      h_sim = 0.0001,
                                      theta_true = theta_0,
                                      theta_start = theta_init,
                                      maxit = 1000,
                                      seed = 1) {
  set.seed(seed)
  results <- vector("list", n_sim * length(h_values))
  row_id <- 1

  for (sim_id in seq_len(n_sim)) {
    full_path <- simulate_student_kramers(
      theta = theta_true,
      T = T,
      h_sim = h_sim
    )

    for (h in h_values) {
      observed_path <- subsample_path(full_path, h = h, h_sim = h_sim)
      fit <- fit_one_simulation(
        path = observed_path,
        h = h,
        theta_start = theta_start,
        maxit = maxit
      )

      results[[row_id]] <- data.frame(
        sim_id = sim_id,
        h = h,
        parameter = parameter_names,
        theta_true = unname(theta_true),
        theta_hat = unname(fit$theta_hat),
        normalized_error = unname((fit$theta_hat - theta_true) / theta_true),
        minimum = fit$minimum,
        loglikelihood = fit$loglikelihood,
        converged = fit$converged,
        convergence = fit$convergence,
        theta_valid = fit$theta_valid,
        function_evaluations = unname(fit$function_evaluations),
        failed = fit$failed
      )

      row_id <- row_id + 1
    }
  }

  bind_rows(results)
}

summarise_simulation_failures <- function(sim_results) {
  sim_results %>%
    distinct(sim_id, h, failed, converged, theta_valid, convergence) %>%
    group_by(h) %>%
    summarise(
      simulations = n(),
      failed = sum(failed),
      converged = sum(converged),
      valid_theta = sum(theta_valid),
      failure_rate = failed / simulations,
      .groups = "drop"
    )
}
```

### Run simulations
```{r, eval=FALSE}
student_kramers_results <- run_student_kramers_study(
  n_sim = 1000,
  h_values = c(0.01, 0.02),
  T = 50,
  h_sim = 0.0001,
  theta_true = theta_0,
  theta_start = theta_init,
  maxit = 1000,
  seed = 1
)

failure_summary <- summarise_simulation_failures(student_kramers_results)
failure_summary
```

### Figure 4 style plot
```{r, eval=FALSE}
plot_data <- student_kramers_results %>%
  filter(!failed)

ggplot(plot_data, aes(x = normalized_error)) +
  geom_density(fill = "steelblue", alpha = 0.35, color = "steelblue4") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_grid(h ~ parameter, scales = "free_y") +
  labs(
    x = expression((hat(theta)[N] - theta[0]) / theta[0]),
    y = "Density"
  ) +
  theme_bw()
```
